They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-14 paper2/paper2.tex
rendered p2

\begin{abstract}
My research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). I find evidence in CoD
supprting the hypothesis that round win probability can be modeled as
constant across the series, although not at a naive 50\%. Second, I
examine post-streak round win probability, in search of evidence
positive recency (the ``hot hand'' fallacy) or negative recency (the
``gambler's falacy''). I find that teams perform significantly worse
than expected after streaks of 2, 3, and 4 wins when series end up going
to 9, 10, or 11 (maximum) rounds, suggesting the presence of negative
recency.
\end{abstract}

@@

Call of Duty (CoD), first released in 2003, is one of the most popular
first-person shooter (FPS) video game franchises of all-time. The most
popular mode in the competitive scene is ``Search and Destroy'' (SnD),
which bears resemblance to ``Bomb Defusal'' in Counter-Strike and
``Plant/Defuse'' in Valorant, two other FPS games played in professional
leagues. SnD is one-sided game mode in which one team, the offensive
side, tries to destroy one of two designated bomb sites on the map.

In professional CoD, a team must win six rounds of SnD to win the
match.\footnote{A maximum of 11 even rounds can be played. There is no
  ``sudden death'' or ``win by two'' rule like there are for SnD
  equivalent in professional Counter-Strike and Valorant matches.} A
round can end in one of fmy ways:

@@

Teams take turns playing offense and defense every round.

@@

CoD has roughly gone through three eras of professional gaming: (1)
Major League Gaming (MLG) tmynaments prior to 2016; (2) the CoD World
League (CWL), initiated in 2016; and (3) the 12-franchise CoD League
(CDL), running since 2020 and completing three year-long ``seasons''
completed as of August 2022.\footnote{CoD is fairly unique compared to
  other esports in that it runs on an annual lifecycle (released coming
  in the late fall), where a new game is published every year under the
  same title. Each new game bears resemblance to past ones, often
  introducing relatively small variations (``improvements'') to
  graphics, game modes, and other facets of gameplay. During the CDL
  era, the games released have been Modern Warfare (2020), Cold War
  (2021) and Vanguard (2022).} The data set consists of all SnD matches
played in tmynaments and qualifiers during the CDL era, totaling 7,792
rounds across 852 series. Data was collected in spreadsheets by
community member ``IOUTurtle''.\footnote{Data:
  \url{https://linktr.ee/CDLArchive}. Author:
  \url{https://twitter.com/IOUTurtle}}

@@

There have been only a handful of studies of the distribution of games
played in a series, most of which assume a constant probability \(p\) of
a given team winning a game in the series, regardless of the series
state. (Mosteller 1952) observed that the American League had dominated
the National League in Major League Baseball's (MLB) World Series
matchups, implying that matchups should not modeled with \(p = 0.5\).
Mosteller proposed three approaches for identifying the optimal constant
probability value of the stronger team in the World Series, finding
\(p \approx 0.65\). in each case: (1) a ``method of moments'' approach
in which one solves for \(p\). from the empirical average number of
games won by the loser of the series; (2) maximizing the likelihood that
the sample would have been drawn from a population in which the
probability of a team winning a game is constant across the series, and
(3) minimizing the chi-square goodness of fit estimate for \(p\).

(Chance 2020) re-examines the constant probability notion in Major
League Baseball's World Series (1923--2018), the National Basketball
Association's Finals (1951--2018), and the National Hockey League's
Stanley Cup (1939--2018). Chance applies finds strong evidence against
the null hypothesis of \(p = 0.5\) in the MLB and NHL championship
series when applying Mosteller's first and second methods.\footnote{Chance
  goes on to outline a conditional probability framework (likelihood of
  winning a game given the series state) which can exactly explain the
  distribution of the number of games played.} Chance's work is closely
related to mys and, in fact, provides a guide for the first part of my
investigation.

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and ``hot hand fallacy''
(positive recency). Per (Ayton and Fischer 2004), negative recency is
``the belief that, for random events, runs of a particular outcome
\ldots{} will be balanced by a tendency for the opposite outcome'',
while positive recency is the expectation of observing future results
that match recent results.

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. (Gilovich, Vallone, and Tversky
1985) do not find evidence for the hot hand phenomenon. Recently ,
(Miller and Sanjurjo 2018) refute (Gilovich, Vallone, and Tversky
1985)'s conclusions, finding mathematical evidence that seems to support
negative recency. Specifically, they find that a ``bias exists in a
common measure of the conditional dependence of present outcomes on
streaks of past outcomes in sequential data'' (streak selection bias)
that imply that, under i.i.d. conditions, ``the proportion of successes
among the outcomes that immediately follow a streak of consecutive
successes is expected to be strictly less than the underlying
(conditional) probability of success''. I agree with Miller's findings,
accounting for the streak selection bias in my study of momentum.

@@

\[
q = 1 - p, s_1 = \frac{s - 1}{2}, s_2 = \frac{s + 1}{2}, s_2^{'} = 1 - s_2
\]

@@

Table \ref{tbl:cod-prob-series-lasting-i-rounds} shows the expected
probabilities for \(s = 11\) under the naive assumption that
\(p_0 = 0.5\) for a given round, along with the observed frequencies
observed in CoD SnD (\(P_O(i)\), \(N_O(i)\))).

\begin{longtable}{rrrr}
  \caption{The probabilities that a best-of-11 series lasts $i$ rounds ($P_E(i)$, where $i = {6, 7, 8, 9, 10, 11}$) under the assumption that each team has a 50% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(i)$ and as a proportion $P_O(i)$ of all series.}\label{tbl:cod-prob-series-lasting-i-rounds} \\
  \toprule
  &  & \multicolumn{2}{c}{Observed} \\ 
  \cmidrule(lr){3-4}
  Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
  \midrule
  6 & $3.1\%$ & $4.7\%$ & 40 \\ 
  7 & $9.4\%$ & $11.9\%$ & 101 \\ 
  8 & $16.4\%$ & $16.5\%$ & 141 \\ 
  9 & $21.9\%$ & $21.7\%$ & 185 \\ 
  10 & $24.6\%$ & $21.5\%$ & 183 \\ 
  11 & $24.6\%$ & $23.7\%$ & 202 \\ 
  \bottomrule
\end{longtable}

The probabilities that a best-of-11 series lasts \(i\) rounds (\(P(i)\),
where \(i \in R = [6, 7, 8, 9, 10, 11]\)) under the assumption that each
team has a 50\% probability (\(p_0 = 0.5\)) of winning each game
\{\#tbl-prob-series-lasting-i-rounds\} along with the observed
frequencies of series lasting \(i\) rounds, expressed as a percentage
\(P_O(i)\) and as a count (\(N_O\))

@@

Table \ref{tbl:mosteller-methods-results} shows the alternate values for
the constant round win probability \(p\) for winning a given round in a
CoD SnD that I find when applying the three methods suggested by
Mosteller (1952). Each is approximately or equal to 0.575, and, when
applying Equation \ref{eq:chi-squ-test}, each results in a \(\chi^2\)
value for which I cannot reject the constant probability null
hypothesis.

@@

Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new
\(P_E(i)\) when re-applying Equation \ref{eq:series-length} for each the
maximum likelihood estimate \(p_2 = 0.575\).\footnote{The method of
  moments and minimum \(\chi^2\) estimates for \(p\) are omitted simply
  because the results would be nearly identical to those for the maximum
  likelihood estimate of \(p\) (since they are all \(\approx 0.575\).} I
observe that \(P_E(i)\) is notably larger for \(p_{2}\) when
\(i \in [6, 7]\), more closely matching \(P_O(i)\). \(p_{2}\) is also
closer to \(P_O(i)\) for \(i \in [9, 10]\), although not for
\(i \in [8, 11]\).

@@

So I conclude that the constant round win probability assumption can be
valid in CoD SnD series with the appropriate choice of \(p\)
(\(\approx 0.575\)).

>>>>

\begin{abstract}
My research investigates patterns in round win percentages in
professional Search and Destroy (SnD) matches of the popular
first-person shooter game Call of Duty (CoD). First, I find evidence
supprting the hypothesis that round win probability can be modeled as a
constant across the series, although not at the naive 50\%. Second, I
examine post-streak round win probability, in search of evidence
positive recency (the ``hot hand'' fallacy) or negative recency (the
``gambler's falacy''). I find that teams perform significantly worse
than expected after streaks of 2, 3, and 4 wins when series end up going
to 9, 10, or 11 (maximum) rounds, suggesting the presence of negative
recency.
\end{abstract}

@@

Call of Duty (CoD), first released in 2003, is one of the most popular
first-person shooter (FPS) video game franchises of all-time. The most
popular mode in the competitive scene is ``Search and Destroy''
(SnD).\footnote{SnD bears resemblance to ``Bomb Defusal'' in
  Counter-Strike and ``Plant/Defuse'' in Valorant, two other FPS games
  played in more popular professional leagues.} SnD is a one-sided game
mode in which one team, the offensive side, tries to destroy one of two
designated bomb sites on the map.

In professional CoD SnD, a team take turns playing offense and defense
every round. They must win six rounds to win the series.\footnote{A
  maximum of 11 even rounds can be played. There is no ``sudden death''
  or ``win by two'' rule like there are for SnD equivalent in
  professional Counter-Strike and Valorant matches.} A round can end in
one of five ways:

@@

@@

CoD has roughly gone through three eras of professional gaming: (1)
Major League Gaming (MLG) tournaments prior to 2016; (2) the CoD World
League (CWL), initiated in 2016; and (3) the 12-franchise CoD League
(CDL), operating since 2020. The CDL has completed three year-long
``seasons'' as of August 2022.\footnote{CoD is fairly unique compared to
  other esports in that it runs on an annual lifecycle (released coming
  in the late fall), where a new game is published every year under the
  same title. Each new game bears resemblance to past ones, often
  introducing relatively small variations (``improvements'') to
  graphics, game modes, and other facets of gameplay. During the CDL
  era, the games released have been Modern Warfare (2020), Cold War
  (2021) and Vanguard (2022).}

The data set consists of all SnD matches played in tournanaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852 series.
Data was collected in spreadsheets by community member
``IOUTurtle''.\footnote{Data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}}

@@

There have been a handful of studies of the distribution of games played
in a series of a professional sport. Most assume a constant probability
\(p\) of a given team winning a game in the series, regardless of the
series state. Mosteller (1952) observed that the American League had
dominated the National League in Major League Baseball's (MLB) World
Series matchups, implying that matchups should not modeled with
\(p = 0.5\). Mosteller proposed three approaches for identifying the
optimal constant probability value of the stronger team in the World
Series, finding \(p \approx 0.65\). in each case: (1) solving for \(p\)
from the empirical average number of games won by the loser of the
series, which he called the ``method of moments'' approach; (2)
maximizing the likelihood that the sample would have been drawn from a
population in which the probability of a team winning a game is constant
across the series (i.e.~maximum likelihood), and (3) minimizing the
chi-square goodness of fit statistic for \(p\).

Chance (2020) re-examines the constant probability notion in Major
League Baseball's World Series (1923--2018), the National Basketball
Association's Finals (1951--2018), and the National Hockey League's
Stanley Cup (1939--2018). Chance finds strong evidence against the null
hypothesis of \(p = 0.5\) in the MLB and NHL championship series when
applying Mosteller's first and second methods.\footnote{Chance goes on
  to outline a conditional probability framework (likelihood of winning
  a game given the series state) which can exactly explain the
  distribution of the number of games played.} Chance's work is closely
related to mine, and, in fact, provides a guide for the first part of my
investigation.

Momentum, one of most discussed topics in sports analytics, goes
hand-in-hand with a discussion of the nature of series
outcomes.\footnote{We often use use ``streaks'' and momentum
  interchangeably, but as (Steeger, Dulin, and Gonzalez 2021) note,
  momentum implies dependence between events, whereas streaking does
  not.} Two opposing fallacies are observed in the context of momentum:
the ``gambler's fallacy'' (negative recency) and ``hot hand fallacy''
(positive recency). Per Ayton et al. (2004), negative recency is ``the
belief that, for random events, runs of a particular outcome \ldots{}
will be balanced by a tendency for the opposite outcome'', while
positive recency is the expectation of observing future results that
match recent results.

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985) do not
find evidence for the hot hand phenomenon. Recently, Miller et al.
(2018) refuted the conclusions of Gilovich et al.~, finding mathematical
evidence that seems to support negative recency. Specifically, they find
that a ``bias exists in a common measure of the conditional dependence
of present outcomes on streaks of past outcomes in sequential data''
(streak selection bias) that imply that, under i.i.d. conditions, ``the
proportion of successes among the outcomes that immediately follow a
streak of consecutive successes is expected to be strictly less than the
underlying (conditional) probability of success''. I agree with Miller's
findings, accounting for the streak selection bias in my study of
momentum.

@@

\[
q = 1 - p, s_1 = \frac{s_1 - 1}{2}, s_2 = \frac{s_1 + 1}{2}, s_2^{'} = 1 - s_2
\]

@@

Table \ref{tbl:cod-prob-series-lasting-i-rounds} shows the expected
probabilities for \(s = 11\) under the naive assumption that
\(p_0 = 0.5\) for a given round, along with the observed proportions
\(P_O(i)\) in CoD SnD.

\begin{longtable}{rrrr}
  \caption{The probabilities that a best-of-11 series lasts $i$ rounds ($P_E(i)$, where $i \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(i)$ and as a proportion $P_O(i)$ of all series.}\label{tbl:cod-prob-series-lasting-i-rounds} \\
  \toprule
  &  & \multicolumn{2}{c}{Observed} \\ 
  \cmidrule(lr){3-4}
  Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
  \midrule
  6 & $3.1\%$ & $4.7\%$ & 40 \\ 
  7 & $9.4\%$ & $11.9\%$ & 101 \\ 
  8 & $16.4\%$ & $16.5\%$ & 141 \\ 
  9 & $21.9\%$ & $21.7\%$ & 185 \\ 
  10 & $24.6\%$ & $21.5\%$ & 183 \\ 
  11 & $24.6\%$ & $23.7\%$ & 202 \\ 
  \bottomrule
\end{longtable}

@@

Table \ref{tbl:mosteller-methods-results} shows the alternate values for
the constant round win probability \(p\) that I find when applying the
three methods suggested by Mosteller (1952). Each is approximately or
equal to 0.575, and, when applying Equation \ref{eq:chi-squ-test}, each
results in a \(\chi^2\) value for which I cannot reject the constant
probability null hypothesis.

@@

Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new
\(P_E(i)\) when re-applying Equation \ref{eq:series-length} for the
maximum likelihood estimate \(p_2 = 0.575\).\footnote{The method of
  moments and minimum \(\chi^2\) estimates for \(p\) are omitted simply
  because the results would be nearly identical to those for the maximum
  likelihood estimate of \(p\) (since they are all \(\approx 0.575\).} I
observe that \(P_E(i)\) is notably larger for \(p_{2}\) when
\(i \in [6, 7]\), more closely matching \(P_O(i)\). \(p_{2}\) is also
closer to \(P_O(i)\) for \(i \in [9, 10]\), although not for
\(i \in [8, 11]\).

@@

It is fair to conclude that the constant round win probability
assumption can be valid in CoD SnD series with the appropriate choice of
\(p\) (\(\approx 0.575\)).
